Arrows for Parallel Computations

### 2017-07-11 paper/content/parallelHaskells.tex
minor

\begin{code}
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = let bs = zipWith ($) fs as 
                 in (bs `using` parList rdeepseq) `pseq` bs
\end{code} %$

In the above definition of |parEvalN| we just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator. We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evalutes to normal form. To ensure that programs that use |parEvalN| have the correct evaluation order, we annotate the computation with |pseq :: a -> b -> b| which forces the compiler to not reorder multiple |parEvalN| computations. This is particularly necessary in circular communication topologies like in the |torus| or |ring| (Chap.~\ref{sec:topology-skeletons}), where a wrong execution order would result in deadlock scenarios when executed without |pseq|.
\olcomment{CAREFUL! is |pseq| REALLY about multiple instances or \emph{merely} order of execution?}
%
Fig.~\ref{fig:parEvalNMulticoreImg} shows a visual representation of this code.

% \begin{figure}[h]
% \begin{code}
% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
% parEvalN fs as = let bs = zipWith ($) fs as 
%                  in (bs `using` parList rdeepseq) `pseq` bs
% \end{code}
% \caption{Multicore version of |parEvalN|.}
% \label{fig:parEvalNMulticore}
% \end{figure}

@@

The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| Monad with |get|. This results in |Par [b]|. We execute this process with |runPar| to finally get |[b]|.
A graphical representation can be found in Fig.~\ref{fig:parEvalNParMonadImg}.

>>>>

\begin{code}
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = let bs = zipWith ($) fs as 
                 in bs `using` parList rdeepseq
\end{code} %$

In the above definition of |parEvalN| we just apply the list of functions |[a -> b]| to the list of inputs |[a]| by zipping them with the application operator |$|. % $
We then evaluate this lazy list |[b]| according to a |Strategy [b]| with the |using :: a -> Strategy a -> a| operator. We construct this strategy with |parList :: Strategy a -> Strategy [a]| and |rdeepseq :: NFData a => Strategy a| where the latter is a strategy which evalutes to normal form.
%
Fig.~\ref{fig:parEvalNMulticoreImg} shows a visual representation of this code.

% \begin{figure}[h]
% \begin{code}
% parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
% parEvalN fs as = let bs = zipWith ($) fs as 
%                  in (bs `using` parList rdeepseq)
% \end{code}
% \caption{Multicore version of |parEvalN|.}
% \label{fig:parEvalNMulticore}
% \end{figure}

@@

The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| Monad with |get|. This results in |Par [b]|. We execute this process with |runPar| to finally get |[b]|.
Fig.~\ref{fig:parEvalNParMonadImg} shows a graphical representation.

### 2017-07-11 paper/content/parrows.tex

\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
The Multicore Haskell implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| and |pseq :: a -> b -> b| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a| but with flipped parameters.
\begin{figure}[t]
\begin{code}
data Conf a = Conf (Strategy a)

instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}% $ %% formatting
\caption{Fully evaluating |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this. We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|.
% \begin{figure}[h]
% \caption{Definition of Conf a}
% \label{fig:confa}
% \end{figure}
% We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances. To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[t]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}
\caption{Configurable |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsubsection{|Par| Monad}
\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
The |Par| Monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| Monad's |spawnP :: NFData a => a -> Par (IVar a)| function. The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| Monad.

>>>>

\subsubsection{Multicore Haskell} \label{sec:parrows:multicore}
The Multicore Haskell implementation of |ArrowParallel| is implemented in a straightforward manner by using |listApp| (Appendix~\ref{utilfns}) combined with the |withStrategy :: Strategy a -> a -> a| combinators from Multicore Haskell, where |withStrategy| is the same as |using :: a -> Strategy a -> a|, but with flipped parameters.
\begin{figure}[t]
\begin{code}
data Conf a = Conf (Strategy a)

instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq))
\end{code}% $ %% formatting
\caption{Fully evaluating |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this. We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|.
% \begin{figure}[h]
% \caption{Definition of Conf a}
% \label{fig:confa}
% \end{figure}
% We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances. To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[t]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}
\caption{Configurable |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsubsection{|Par| Monad}
\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
The |Par| Monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| Monad's |spawnP :: NFData a => a -> Par (IVar a)| function. The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| Monad.
